Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-02-14 paper.tex
One more revision in the proof of DV

We will need one more root-theoretic lemma before we proceed with the proof of~\cref{thm:DennisVaserstein}.
Let $\Phi$, $r$ and $s$ be as in~\cref{table:dv-reps}.
Denote by $O$ the subset of roots of $\Sigma_s^-$ lying in the orbit $W(\Delta_s) \cdot (\Sigma^-_s\cap \Delta_r)$.
\begin{lemma} \label{lemma:root-lemma}
Suppose that $\Phi \neq \rC_\ell$. The following alternative holds: either $O$ coincides with $\Sigma_s^-$ or
$\Phi=\rB_\ell, \rF_4$ and $\Sigma_s^- \setminus O$ consists of the sole short root $\gamma$, which can be presented as a sum 
 $\gamma = \alpha + \beta$ for some short root $\alpha \in \Phi^+$ and some long root $\beta \in O$ satisfying $\beta + 2\alpha \in O$. 
\end{lemma}
\begin{proof}
From \cref{lemma:abs} it follows that $O$ does not coincide with $\Sigma^-_s$ only in the following two cases.
 \begin{enumerate} 
 \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
 Only the sole short root of $\Sigma_1^-$ is not contained in $O$, denote it by $\gamma$. Set $\alpha = \alpha_2 + \ldots + \alpha_\ell,\ \beta = -\alpha_1 - 2\alpha_2 - \ldots -2\alpha_\ell.$
 It is clear that $\alpha \in \Phi^+$, $\beta, \beta + 2\alpha \in O$.
 \item \textit{Case $\Phi = \rF_4$, $s=4$, $r=1$.} There are three possibilities for the $s$-shape and length of $\alpha\in \Sigma_4^-$: 
 \begin{itemize}
  \item $\shape(\{\alpha_4\}, \alpha) = 1$, $\alpha$ is short (there are 8 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is long (there are 6 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is short (there is only one such root, denote it $\gamma$).
 \end{itemize}
 Since $\Delta_r \cap \Sigma_s^-$ contains roots of the first two types, only $\gamma$ is not contained in $O$.
 Clearly, $\gamma = -\alpha_1 -2\alpha_2-3\alpha_3-2\alpha_4 = \alpha + \beta$ for $\alpha = \alpha_2 + \alpha_3\in \Phi^+,$ $\beta = -\alpha_{{\max}} \in O$. \qedhere
\end{enumerate}
\end{proof}

\begin{proof} [Proof of \cref{thm:DennisVaserstein}]
From Lemmas~\ref{lemma:Stein_reduction} and~\ref{lemma:dv-normal} it follows that $N(A_{rs})$ contains $\EP_s(R)$.
Thus, by~\cref{lem:LN} we obtain that $\EP_s(R, I) \subseteq L(A_{rs})$.

Our next goal is to verify the inclusion $\U(\Sigma_s^-, I) \subseteq L(A_{rs})$.
Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N(A_{rs})$ and 
 $X_{\alpha}(I) \subseteq L(A_{rs})$ for $\alpha \in \Sigma^-_s\cap \Delta_r$, 
we obtain from~\cref{lem:LN} and \cref{item-trans2} that $L(A_{rs})$ contains root subgroups $X_\alpha(I)$ for $\alpha \in O = W(\Delta_s) \cdot (\Sigma^-_s\cap \Delta_r)$.
Thus, by~\cref{lemma:root-lemma} we are left to consider two cases when $O$ does not coincide with $\Sigma^-_s$. 
\begin{enumerate} 
  \item \textit{Case $\Phi = \rB_\ell, \rF_4$.} Let $\alpha, \beta, \gamma$ be as in~\cref{lemma:root-lemma}.
  Since $X_{\beta}(I), X_{\beta+2\alpha}(I) \subseteq L(A_{rs})$ and $X_{\alpha}(R) \subseteq N(A_{rs})$ we obtain the required inclusion
   $X_{\gamma}(I) \subseteq L(A_{rs})$ using Chevalley commutator formula:
   \begin{equation} \nonumber x_{\gamma}(\pm ab) = [x_{\beta}(a),\, x_{\alpha}(b)] \cdot x_{\beta + 2\alpha}(ab^2). \end{equation}
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  In this case we need to invoke stability assumptions one more time.
  From~\cref{lemma:DVST} below we obtain that $N(A_{rs})$ contains $X_{-\alpha_1}(R)$ and therefore coincides with $\E(\Phi, R)$.
  The required inclusion now follows~\cref{item-trans1}.
\end{enumerate}
Thus, we have shown that $\mathcal{X}\subseteq L(A_{rs})$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by the subset $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L(A_{rs})^{N(A_{rs})} \subseteq L(A_{rs})$ we conclude that 
 $\E(\Phi, R, I) \subseteq L(A_{rs})$, which completes the proof.
\end{proof}

\begin{lemma}\label{lemma:DVST}
Assume that $\Phi=\rC_\ell$ and $\sr(I) \leq \ell-1$, then $X_{-\alpha_s}(R) \subseteq N(A_{rs})$.
\end{lemma}
\begin{proof}
Denote by $C$ the set consisting of elements $g \in \EP_s(R, I)$ for which matrix entries $(g_{i, 2})$, $i=2, \ldots, \ell$ form an $I$-unimodular column of height $\ell-1$

Set $A' = \EP_\ell(R, I) \cdot \U(\Sigma_s^- \cap \Sigma_r^-, I) \cdot C$.
Applying \cref{item:asrUnipC} we find for every $g \in \EP_s(R, I)$ an element $x \in \U(\Sigma_r^+ \cap \Delta_s, I)$ such that $xg \in C$.  
Notice that the equality $A_{rs} = A'$ follows from this, indeed, for $g\in \EP_s(R, I)$ one has
\begin{equation*} \EP_r(R, I) \cdot \U(\Sigma_s^- \cap \Sigma_r^-, I) \cdot g \subseteq 
 \EP_r(R, I)x^{-1}  \cdot \U(\Sigma_s^-, I) \cdot xg \subseteq A'. \end{equation*}

By the definition of $C$, for every $g \in C$ one can choose $y \in \U(\Sigma_s^+ \cap \Delta_r, I)$ such that $(y \cdot g)_{\varpi_s, \varpi_s - \alpha_s} = 0$.
Consequently, for every $g\in C$ one has
\[
 \EP_r(R, I) \cdot \U(\Sigma_s^- \cap \Sigma_r^-, I) \cdot g \subseteq \EP_r(R, I) y^{-1} \cdot \U(\Sigma_r^-\cap \Sigma_s^-, I)^{y^{-1}} \cdot y g
  \subseteq \EP_r(R, I) \cdot \U(\Sigma_r^-, I) \cdot y g.
\]
Notice that the matrix entry $(yg)_{\varpi_s, \varpi_s}$ is invertible.
From the choice of $y$ it follows that for every $\xi\in R$ the element $g_1 \coloneqq (yg)^{x_{-\alpha_s}(\xi)}$
satisfies the assumptions of \cref{lemma:Chevalley-Matsumoto} and therefore can be rewritten as $g_1 = uh$ for some $u \in \U(\Sigma_s^-, I)$, $h \in \EP_s(R, I)$.
Consequently, one has
\begin{multline*} {A_{rs}}^{X_{-\alpha_{s}}(R)} \subseteq \EP_r(R, I)^{X_{-\alpha_{s}}(R)} \cdot \U(\Sigma_r^-, I)^{X_{-\alpha_{s}}(R)} \cdot \U(\Sigma_s^-, I) \cdot \EP_s(R, I) \subseteq \\
\subseteq \EP_r(R, I) \cdot \U(\Phi^-, I) \cdot \EP_s(R, I) \subseteq A_{rs}, \end{multline*}
as claimed.
\end{proof}

>>>>

\begin{lemma}\label{lemma:DVST}
Assume that $\Phi=\rC_\ell$, $s=1$, $r=\ell$ and that $\sr(I) \leq \ell-1$. Then $X_{-\alpha_1}(R) \subseteq N(A_{\ell 1})$.
\end{lemma}
\begin{proof}
Denote by $C$ the set consisting of elements $g \in \EP_1(R, I)$ for which matrix entries $(g_{i, 2})$, $i=2, \ldots, \ell$ form an $I$-unimodular column of height $\ell-1$.
Set $A' = \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot C$.
Applying \cref{item:asrUnipC} we can find for every $g \in \EP_1(R, I)$ some element $x \in \U(\Sigma_\ell^+ \cap \Delta_1, I)$ such that $xg \in C$.  
The equality $A_{\ell 1} = A'$ follows from this, indeed, for $g\in \EP_1(R, I)$ one has
\begin{equation*} \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot g \subseteq 
 \EP_\ell(R, I)x^{-1}  \cdot \U(\Sigma_1^-, I) \cdot xg \subseteq A'. \end{equation*}

By the definition of $C$, for every $g \in C$ one can choose $y \in \U(\Sigma_1^+ \cap \Delta_\ell, I)$ such that $(y \cdot g)_{1, 2} = 0$.
Consequently, for every $g\in C$ one has
\[
 \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot g \subseteq \EP_\ell(R, I) y^{-1} \cdot \U(\Sigma_\ell^-\cap \Sigma_1^-, I)^{y^{-1}} \cdot y g
  \subseteq \EP_\ell(R, I) \cdot \U(\Sigma_\ell^-, I) \cdot y g.
\]
Notice that the matrix entry $(yg)_{1, 1}$ is invertible.
From the choice of $y$ it follows that for every $\xi\in R$ the element $g_1 \coloneqq (yg)^{x_{-\alpha_1}(\xi)}$
satisfies the assumptions of \cref{lemma:Chevalley-Matsumoto} and therefore can be rewritten as $g_1 = uh$ for some $u \in \U(\Sigma_1^-, I)$, $h \in \EP_1(R, I)$.
Consequently, one has
\begin{multline*} {A_{\ell 1}}^{X_{-\alpha_{1}}(R)} \subseteq \EP_\ell(R, I)^{X_{-\alpha_1}(R)} \cdot \U(\Sigma_\ell^-, I)^{X_{-\alpha_1}(R)} \cdot \U(\Sigma_1^-, I) \cdot \EP_1(R, I) \subseteq \\
\subseteq \EP_\ell(R, I) \cdot \U(\Phi^-, I) \cdot \EP_1(R, I) \subseteq A_{\ell 1}, \end{multline*} as claimed.
\end{proof}

Now we are all set to prove the main result of this subsection.
\begin{proof} [Proof of \cref{thm:DennisVaserstein}]
From Lemmas~\ref{lemma:Stein_reduction} and~\ref{lemma:dv-normal} it follows that $N(A_{rs})$ contains $\EP_s(R)$.
Thus, by~\cref{lem:LN} we obtain that $\EP_s(R, I) \subseteq L(A_{rs})$.

Our next goal is to verify the inclusion $\U(\Sigma_s^-, I) \subseteq L(A_{rs})$.
Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N(A_{rs})$ and 
 $X_{\alpha}(I) \subseteq L(A_{rs})$ for $\alpha \in \Sigma^-_s\cap \Delta_r$, 
we obtain from~\cref{lem:LN} and \cref{item-trans2} that $L(A_{rs})$ contains root subgroups $X_\alpha(I)$ for $\alpha$ lying in the orbit $O = W(\Delta_s) \cdot (\Sigma^-_s\cap \Delta_r)$.
Thus we are left to consider three cases when $O$ does not coincide with $\Sigma^-_s$. 
\begin{enumerate} 
  \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
 Only the sole short root of $\Sigma_1^-$ is not contained in $O$, denote it by $\gamma$.
 Set $\alpha = \alpha_2 + \ldots + \alpha_\ell,\ \beta = -\alpha_1 - 2\alpha_2 - \ldots -2\alpha_\ell.$
 It is clear that $\alpha \in \Phi^+$, $\beta, \beta + 2\alpha \in O$.
 Since $X_{\beta}(I), X_{\beta+2\alpha}(I) \subseteq L(A_{rs})$ and $X_{\alpha}(R) \subseteq N(A_{rs})$ we obtain the required inclusion $X_{\gamma}(I) \subseteq L(A_{rs})$ using Chevalley commutator formula:
 \begin{equation} \nonumber x_{\gamma}(\pm ab) = [x_{\beta}(a),\, x_{\alpha}(b)] \cdot x_{\beta + 2\alpha}(ab^2). \end{equation}
 \item \textit{Case $\Phi = \rF_4$, $s=4$, $r=1$.}
 There are three possibilities for the $s$-shape and length of $\alpha\in \Sigma_4^-$: 
 \begin{itemize}
  \item $\shape(\{\alpha_4\}, \alpha) = 1$, $\alpha$ is short (there are 8 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is long (there are 6 such roots);
  \item $\shape(\{\alpha_4\}, \alpha) = 2$, $\alpha$ is short (there is only one such root, denote it by $\gamma$).
 \end{itemize}
 Since $\Delta_r \cap \Sigma_s^-$ contains roots of the first two types, only $\gamma$ is not contained in $O$.
 Clearly, $\gamma = -\alpha_1 -2\alpha_2-3\alpha_3-2\alpha_4 = \alpha + \beta$ for $\alpha = \alpha_1 + \alpha_2 + \alpha_3\in \Phi^+,$ $\beta = -\alpha_{{\max}} \in O$.
 Since $\alpha \in \Phi^+$, $\beta, \beta + 2\alpha \in O$ we obtain the inclusion $X_{\gamma}(I) \subseteq L(A_{rs})$ by the same token as in the previous case.
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  In this case we need to invoke stability assumptions one more time.
  From~\cref{lemma:DVST} we obtain that $N(A_{rs})$ contains $X_{-\alpha_1}(R)$ and therefore coincides with $\E(\Phi, R)$.
  The required inclusion now follows~\cref{item-trans1}.
\end{enumerate}
Thus, we have shown that $\mathcal{X}\subseteq L(A_{rs})$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by the subset $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L(A_{rs})^{N(A_{rs})} \subseteq L(A_{rs})$ we conclude that 
 $\E(\Phi, R, I) \subseteq L(A_{rs})$, which completes the proof.
\end{proof}
